Inference in Hidden Markov Models with Explicit State Duration Distributions

### 2011-12-17 tex/edhmm.tex
made reviewer typo edits

The need to accommodate explicit state duration distributions in HMMs has long been recognised. Rabiner \cite{Rabiner89} details the basic approach which expands the state space to include dwell time before applying a slightly modified Baum-Welch algorithm. This approach specifies a maximum state duration, limiting practical application 
to cases with short sequences and dwell times.
This approach, generalised under the name ``segmental hidden Markov models'', includes more general transitions than those Rabiner considered, allowing the next state and duration to be conditioned on the previous state and duration \cite{Gales93}. Efficient approximate inference procedures were developed in the context of speech recognition \cite{Ostendorf96} and evolved into symmetric approaches suitable for practical implementation \cite{Yu2006}.

>>>>

The need to accommodate explicit state duration distributions in HMMs has long been recognised. Rabiner \cite{Rabiner89} details the basic approach which expands the state space to include dwell time before applying a slightly modified Baum-Welch algorithm. This approach specifies a maximum state duration, limiting practical application 
to cases with short sequences and dwell times.
This approach, generalised under the name ``segmental hidden Markov models'', includes more general transitions than those Rabiner considered, allowing the next state and duration to be conditioned on the previous state and duration \cite{Gales93}. Efficient approximate inference procedures were developed in the context of speech recognition \cite{Ostendorf96}, speech synthesis \cite{Zen07}, and evolved into symmetric approaches suitable for practical implementation \cite{Yu2006}.

### 2011-12-17 tex/introduction.tex
made reviewer typo edits

\label{section}
Hidden Markov models (HMMs) are a fundamental tool for data analysis and exploration.  Many variants of the basic HMM have been developed in response to shortcomings in the original HMM formulation \cite{Rabiner89}.  In this paper we address inference in the explicit state duration HMM.  By state duration we mean the amount of time an HMM dwells in a state.  In the standard HMM specification, a state's duration is implicit and, a priori, distributed geometrically.

The explicit state duration HMM (EDHMM) \cite{Rabiner89} was developed to allow explicit parameterization and direct inference of state duration distributions.   EDHMM estimation and inference can be performed using the forward-backward algorithm; specifically if the sequence is short or a tight ``allowable'' duration interval for each state is hard-coded a priori \cite{Yu2006}.   If the sequence is short then forward-backward can be run on a state representation that allows for all possible durations up to the observed sequence length.  If the sequence is long then forward-backward only remains computationally tractable if only transitions between durations that lie within pre-specified allowable intervals are considered.   If the true state durations lie outside those intervals then the resulting model estimates will be incorrect: the learned duration distributions will reflect only what is allowed given the pre-specified duration intervals. % (which may poorly describe the true duration distributions).

>>>>

\label{section}
Hidden Markov models (HMMs) are a fundamental tool for data analysis and exploration.  Many variants of the basic HMM have been developed in response to shortcomings in the original HMM formulation \cite{Rabiner89}.  In this paper we address inference in the explicit state duration HMM (EDHMM).  By state duration we mean the amount of time an HMM dwells in a state.  In the standard HMM specification, a state's duration is implicit and, a priori, distributed geometrically.

The explicit state duration HMM (EDHMM) \cite{Rabiner89}, a member of the Hidden Semi Markov Models \cite{Yu10}, was developed to 
allow explicit parameterization and direct inference of state duration 
distributions.   EDHMM estimation and inference can be performed using the 
forward-backward algorithm; specifically if the sequence is short or a tight 
``allowable'' duration interval for each state is hard-coded a priori 
\cite{Yu2006}.   If the sequence is short then forward-backward can be run 
on a state representation that allows for all possible durations up to the 
observed sequence length.  If the sequence is long then forward-backward 
only remains computationally tractable if only transitions between durations 
that lie within pre-specified allowable intervals are considered.   If the 
true state durations lie outside those intervals then the resulting model 
estimates will be incorrect: the learned duration distributions will reflect 
only what is allowed given the pre-specified duration intervals. % (which may poorly describe the true duration distributions).

### 2011-12-17 tex/results.tex
final update
made reviewer typo edits

Broad, uninformative priors were chosen for the parameters of the duration and observation distributions. The the observation distribution parameters were given a normal-inverse-Wishart (N-IW) prior with parameters $\nu_0 = 2$, $\Lambda_0 = 1$, $\kappa=0.1$ and $\mu_0 = 0$. The rate parameters for all states were given $\mathrm{Gamma}(1, 10^{5})$ priors.

@@

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_Y.pdf}
        \label{fig:exp_2_data}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_X.pdf}
        \label{fig:exp_2_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_means_exp3.pdf}
        \label{fig:posterior_means_3}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates_exp3.pdf}
        \label{fig:posterior_rates_3}
    }
    \caption{Beam sampler results applied to a system with identical observation distributions but differing durations. Observations are shown in a); true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler. Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$. Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}
    \label{fig:experiment2_results}
\end{figure}

>>>>

Broad, uninformative priors were chosen for the parameters of the duration and observation distributions. The observation distribution parameters were given a normal-inverse-Wishart (N-IW) prior with parameters $\nu_0 = 2$, $\Lambda_0 = 1$, $\kappa=0.1$ and $\mu_0 = 0$. The rate parameters for all states were given $\mathrm{Gamma}(1, 10^{5})$ priors.

@@

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_Y.pdf}
        \label{fig:exp_2_data}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_X.pdf}
        \label{fig:exp_2_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_means_exp3.pdf}
        \label{fig:posterior_means_3}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates_exp3.pdf}
        \label{fig:posterior_rates_3}
    }
    \caption{Beam sampler results applied to a system with identical observation distributions but differing durations. Observations are shown in a); true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler. Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$ and the duration distributions have rates $\lambda_1 = 5$, $\lambda_2 = 15$, $\lambda_3 = 20$. Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}
    \label{fig:experiment2_results}
\end{figure}

### 2011-12-17 tex/sampling.tex
my final (I hope) edits
made reviewer typo edits

We can, in theory, use the forward messages from the forward backward algorithm \cite{Rabiner89} to sample the conditional posterior distribution of $\mathcal{X}$ and $\mathcal{D}.$   To do this we treat each state-duration tuple as a single random variable 
%(i.e.~$z_t = \{x_t,d_t\}$).  
(introducing the notation $z_t = \{x_t,d_t\}$).  
Doing so recovers the standard hidden Markov model structure and hence standard forward messages can be used directly.  A forward filtering, backward sampler for $\mathcal{Z} = \{z_1, \ldots, x_T\}$ conditioned on 
%everything else 
all other random variables
requires the classical forward messages:
    \begin{equation}
        \alpha_t(z_t) = 
        \sum_{z_{t-1}}
        p(z_t | z_{t-1}) 
        p(y_t|z_t) 
        \alpha_t(z_{t-1}) 
        \label{eqn:forward recursion}
    \end{equation}
   % \begin{equation}
   % \beta(z_t) = \sum_{z_{t+1}} p(z_{t+1} | z_t) p(y_{t+1}|z_{t+1}) \beta_{t+1}(z_{t+1})
   % \label{eqn:backward recursion}
    % \end{equation}
     where the transition probability can be factorised according to our modelling assumptions:
     \begin{equation}
        p(z_{t} | z_{t-1}) = p(x_t | x_{t-1}, d_{t-1}) p(d_t | d_{t-1}, x_t) p(y_t|x_t).
     \end{equation}

@@

The particular choice of auxiliary variable $u_t$ is important.  We follow  \cite{vanGael2008} in choosing $u_t$ to be conditionally distributed given the current and previous state and duration in the following way (see the graphical model in Figure \ref{fig:aux graphical model}):
\begin{equation}
    \label{eqn:slice}
    p(u_t | z_t, z_{t-1}) = 
    \frac
    {\mathbb{I}(0 < u_t < p(z_t | z_{t-1}))} 
    {p(z_t | z_{t-1})}.
\end{equation}
Given $\mathcal{U}$ it possible to sample the state $\mathcal{X}$ and duration $\mathcal{D}$ conditional posterior. 
%\begin{equation}
%    \hat{\alpha}_t(z_t) = p(z_t, d_t , \mathcal{Y}_1^t, \mathcal{U}_1^t)
%\end{equation}
Using notation $\mathcal{Y}_{t_1}^{t_2} = \{y_{t_1}, y_{t_1+1}, \ldots,y_{t_2}\}$  to indicate sub-ranges of a sequence, the new forward messages we compute are:
\begin{eqnarray}
   \hat{\alpha}_t(z_t) &=& 
   p(z_t, \mathcal{Y}_1^t , \mathcal{U}_1^{t-1})   \label{eqn:scaled forward} \\
   &=& 
   \sum_{z_{t-1}}
   p(z_t, z_{t-1} , \mathcal{Y}_1^t , \mathcal{U}_1^{t-1}) \nonumber \\
   &\propto& 
   \sum_{z_{t-1}}
   p(u_{t-1} | z_t, z_{t-1})
   p(z_t, z_{t-1} , \mathcal{Y}_1^t, \mathcal{U}_1^{t-1}) \nonumber \\
   &=& 
   \sum_{z_{t-1}}
   \mathbb{I}(0 < u_{t-1} < p(z_t | z_{t-1}))
   p(y_t|z_t) \hat{\alpha}_{t-1}(z_{t-1}) \nonumber.
\end{eqnarray}
The indicator function results in non-zero probabilities in the forward message for only those states $z_t$ whose likelihood given $z_{t-1}$ is greater than $u_t$. The beam sampler derives its computational advantage from the fact that the set of $z_t$'s for which this is true is typically small. 
%Hence \eqref{eqn:indicator} can be rewritten as a limitation of the sum over possible state transitions:
%\begin{equation}
%    \hat{\alpha}_t(z_t) = 
%    \begin{cases} 
%            p(y_t|z_t)\sum_{z_{t-1}}\hat{\alpha}_{t-1}(z_{t-1}) & \textrm{if $u_t < p(z_t | z_{t-1})$} \\
%            0 & \textrm{otherwise}
%        \end{cases}    
%    \label{eqn:scaled forward}
%\end{equation}
%which has restricted the calculation to a finite set. 
%Care must be taken to find those transitions $p(z_t | z_{t-1})$ whose likelihood is greater than the auxiliary variable $u_t$, though this is a straightforward task for unimodal distributions.

@@

The backwards sampling step recursively samples a state sequence from the distribution $p(z_t | z_{t+1}, \mathcal{Y}, \mathcal{U})$
%\begin{equation}
%    p(z_t | z_{t+1}, \mathcal{Y}, \mathcal{U}).
%\end{equation}
which can expressed in terms of the forward variable:
\begin{eqnarray}
    p(z_t | z_{t+1}, \mathcal{Y}, \mathcal{U}) &\propto& p(z_{t+1},z_t, \mathcal{Y}, \mathcal{U}) \nonumber \\
    & \propto &  
        p(u_{t} | z_t, z_{t+1})p(z_{t+1}|z_t) \hat{\alpha}_t(z_t)
        %p(z_t, \mathcal{Y}_1^t, \mathcal{U}_1^t)
        %p(\mathcal{Y}_{t+1}^T, \mathcal{U}_{t+2}^T | z_{t+1}) 
        \nonumber\\
    & \propto & 
       \mathbb{I}(0 < u_{t} < p(z_{t+1} | z_t))
        \hat{\alpha}_t(z_t).
        \label{eqn:backward}
\end{eqnarray}

>>>>

We can, in theory, use the forward messages from the forward backward algorithm \cite{Rabiner89} to sample the conditional posterior distribution of $\mathcal{X}$ and $\mathcal{D}.$   To do this we treat each state-duration tuple as a single random variable 
%(i.e.~$z_t = \{x_t,d_t\}$).  
(introducing the notation $z_t = \{x_t,d_t\}$).  
Doing so recovers the standard hidden Markov model structure and hence standard forward messages can be used directly.  A forward filtering, backward sampler for $\mathcal{Z} = \{z_1, \ldots, z_T\}$ conditioned on 
%everything else 
all other random variables
requires the classical forward messages:
    \begin{equation}
        \alpha_t(z_t) = 
        \sum_{z_{t-1}}
        p(z_t | z_{t-1}) 
        p(y_t|z_t) 
        \alpha_t(z_{t-1}) 
        \label{eqn:forward recursion}
    \end{equation}
   % \begin{equation}
   % \beta(z_t) = \sum_{z_{t+1}} p(z_{t+1} | z_t) p(y_{t+1}|z_{t+1}) \beta_{t+1}(z_{t+1})
   % \label{eqn:backward recursion}
    % \end{equation}
     where the transition probability can be factorised according to our modelling assumptions:
     \begin{equation}
        p(z_{t} | z_{t-1}) = p(x_t | x_{t-1}, d_{t-1}) p(d_t | d_{t-1}, x_t).
     \end{equation}

@@

The particular choice of auxiliary variable $u_t$ is important.  We follow  \cite{vanGael2008} in choosing $u_t$ to be conditionally distributed given the current and previous state and duration in the following way (see the graphical model in Figure \ref{fig:aux graphical model}):
\begin{equation}
    \label{eqn:slice}
    p(u_t | z_t, z_{t-1}) = 
    \frac
    {\mathbb{I}(0 < u_t < p(z_t | z_{t-1}))} 
    {p(z_t | z_{t-1})}.
\end{equation}
where $\mathbb{I}(\cdot)$ returns one if its operand is true and zero otherwise. Given $\mathcal{U}$ it is possible to sample the state $\mathcal{X}$ and duration $\mathcal{D}$ conditional posterior. 
%\begin{equation}
%    \hat{\alpha}_t(z_t) = p(z_t, d_t , \mathcal{Y}_1^t, \mathcal{U}_1^t)
%\end{equation}
Using notation $\mathcal{Y}_{t_1}^{t_2} = \{y_{t_1}, y_{t_1+1}, \ldots,y_{t_2}\}$  to indicate sub-ranges of a sequence, the new forward messages we compute are:
\begin{eqnarray}
   \hat{\alpha}_t(z_t) &=& 
   p(z_t, \mathcal{Y}_1^t , \mathcal{U}_1^{t})   \label{eqn:scaled forward} \\
   &=& 
   \sum_{z_{t-1}}
   p(z_t, z_{t-1} , \mathcal{Y}_1^t , \mathcal{U}_1^{t}) \nonumber \\
   &\propto& 
   \sum_{z_{t-1}}
   p(u_{t} | z_t, z_{t-1})
   p(z_t, z_{t-1} , \mathcal{Y}_1^t, \mathcal{U}_1^{t-1}) \nonumber \\
   &=& 
   \sum_{z_{t-1}}
   \mathbb{I}(0 < u_{t} < p(z_t | z_{t-1}))
   p(y_t|z_t) \hat{\alpha}_{t-1}(z_{t-1}) \nonumber.
\end{eqnarray}
The indicator function $\mathbb{I}$ results in non-zero probabilities in the forward message for only those states $z_t$ whose likelihood given $z_{t-1}$ is greater than $u_t$. The beam sampler derives its computational advantage from the fact that the set of $z_t$'s for which this is true is typically small. 
%Hence \eqref{eqn:indicator} can be rewritten as a limitation of the sum over possible state transitions:
%\begin{equation}
%    \hat{\alpha}_t(z_t) = 
%    \begin{cases} 
%            p(y_t|z_t)\sum_{z_{t-1}}\hat{\alpha}_{t-1}(z_{t-1}) & \textrm{if $u_t < p(z_t | z_{t-1})$} \\
%            0 & \textrm{otherwise}
%        \end{cases}    
%    \label{eqn:scaled forward}
%\end{equation}
%which has restricted the calculation to a finite set. 
%Care must be taken to find those transitions $p(z_t | z_{t-1})$ whose likelihood is greater than the auxiliary variable $u_t$, though this is a straightforward task for unimodal distributions.

@@

The backwards sampling step recursively samples a state sequence from the distribution $p(z_{t-1} | z_{t}, \mathcal{Y}, \mathcal{U})$
%\begin{equation}
%    p(z_t | z_{t+1}, \mathcal{Y}, \mathcal{U}).
%\end{equation}
which can expressed in terms of the forward variable:
\begin{eqnarray}
    p(z_{t-1} | z_{t}, \mathcal{Y}, \mathcal{U}) &\propto& p(z_{t},z_{t-1}, \mathcal{Y}, \mathcal{U})  \label{eqn:backward} \\
    & \propto &  
        p(u_{t} | z_t, z_{t-1})p(z_{t}|z_{t-1}) \hat{\alpha}_{t-1}(z_{t-1})
        %p(z_t, \mathcal{Y}_1^t, \mathcal{U}_1^t)
        %p(\mathcal{Y}_{t+1}^T, \mathcal{U}_{t+2}^T | z_{t+1}) 
        \nonumber\\
    & \propto & 
       \mathbb{I}(0 < u_{t} < p(z_{t} | z_{t-1}))
        \hat{\alpha}_{t-1}(z_{t-1}).\nonumber
\end{eqnarray}
